Quantifying uncertainty in Pareto lake area estimates

### 2021-09-07 manuscript/manuscript.tex
minor grammar fixes

\author{ \href{https://orcid.org/0000-0002-5924-2464}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}J.~Stachelek}\thanks{\href{https://jsta.rbind.io}{https://jsta.rbind.io}} \\
	Center for Limnology\\
	University of Wisconsin-Madison\\
	Madison, WI 53706 \\
	\texttt{stachelek@wisc.edu} \\
}

@@

\section{Introduction}
Size is a critical factor determining the contribution of lakes to global biogeochemical cycling and existing evidence suggests that lakes are key components of overall carbon burial, sequestration, and greenhouse gas emissions \citep{delsontroGreenhouseGasEmissions2018, kellerGlobalCarbonBudget2021}. The calculation of such fluxes requires two terms 1) an areal flux rate and 2) the distribution of total lake area. As with all estimates at broad spatial scales, uncertainty estimates are of great interest both because they allow for intercomparisons among similar lake studies and because they allow for comparisons between lakes and other ecosystems.

The techniques typically used to determine uncertainty in areal flux rates and the techniques typically used to determine uncertainty in total lake area distributions differ greatly. In the former case, uncertainties can typically be calculated in straightforward manner from literature compilations \citep{delsontroGreenhouseGasEmissions2018, kellerGlobalCarbonBudget2021}. In the latter case, uncertainty in total lake area distributions is challenged by the fact that no existing database is a complete census of all lakes \citep{messagerEstimatingVolumeAge2016} and the distribution of individual lake areas spans a range of approximately 7 orders of magnitude. The largest lakes are so big they could otherwise be classified as inland seas (> $10^4$ $km^2$) while the smallest are barely larger than a regulation sized soccer field (> $10^{-3}$ $km^2$).
A consequence of these challenges is that uncertainty varies within the distribution as a function of lake size. The area of the largest lakes is known exactly while the area of the smallest lakes is incomplete below a certain unknown threshold. This can occur either because lakes are too small to be resolved from remote sensing products or because of limited ground surveying effort. The omission or "censoring" of small lakes occurs because we know that small lakes exist but below a certain threshold we have limited knowledge of their exact areas \citep{hamiltonEstimationFractalDimension1992}.

Estimating total lake area from a sample of lakes requires a conceptual model of how lakes are formed (i.e. the data generating process). Typically, lake areas are treated as arising from a fractal generating process due to the fact that landform topography, which determines the placement of lakes, can itself be treated as a fractal generating process. Indeed, many other geomorphological phenomena that are dependent on landform topography such as coastline length are often well-described by fractal generating processes \citep{newman_power_2005}. A challenge in modelling such data generating processes for lake areas is large lakes likely follow a different data generating process than that of the smallest lakes. Whereas small lakes are constrained by landform topography, large lakes are essentially unconstrained by local landform topography and are instead constrained by the placement and arrangement of continents \citep{goodchildLakesFractalSurfaces1988}. As large lakes become even larger, they have a greater probability of intersecting a continent edge and becoming an estuary or embayment rather than a lake. Consquently, lake databases are said to be truncated on large lakes because we know that large lakes are essentially fixed in space and cannot occur in any given location \citep{hamiltonEstimationFractalDimension1992}.

From the preceding discussion it is clear that estimating total lake area requires a method of dealing both with the fact that lake databases are 1) truncated at large lakes and 2) censored at small lakes. Previous studies estimating global lake area have essentially not dealt with the first issue but instead modified their estimation process to limit its effect on the results (see methods section). They have dealt with the second issue by specifying an ad-hoc cutoff value below which empirically determined lake areas are discarded and subsequently back calculated. Hereafter, I refer to this strategy as the cutoff method. In the present study, I explore the effects of using the typical frequentist approach for estimating total lake area (i.e. the cutoff method) via a simulation study. In addition, I demonstrate an approach to produce uncertainty estimates of total lake area in a Bayesian framework. I evaluate the effects of the cutoff method on a simulated dataset because it is not sensitive to other potential confounding factors such as heterogeneity of survey effort or data precision.

@@

Lake areas are typically treated as arising from a scale-invariant fractal generating process \citep{winslowDoesLakeSize2015, downingGlobalAbundanceSize2006, mcdonald_regional_2012, goodchildLakesFractalSurfaces1988, hamiltonEstimationFractalDimension1992}. Essentially, this means that the number of lakes in one size class is proportional to the number of lakes in the preceding size class irrespective of their magnitudes. The numerical form describing such a process is a power-law function. One of the statistical tools often used to model data that follow a power-law function is the Pareto distribution which has a probability density function (pdf) of:

@@

The reason for using the ccdf is two-fold. First, it stabilizes model estimates in the lower tail of the distribution \citep{newman_power_2005}. This can be seen from the simulated data in Figure \ref{fig:pareto_demo} where the Pareto pdf contains a lot of noise in the tail but the ccdf appears smoothed. The smoothing of the tail is also desirable because it functions as a way of dealing with the truncated nature of lake databases. The second reason for using the ccdf is that it provides a computational shortcut for estimating the Pareto shape parameter $a$ because it is numerically equivalent to the slope of the ccdf in log-log space \citep{downingGlobalAbundanceSize2006}.

@@

For evaluation purposes, I generated a simulated dataset of 10,000 lake areas following the Pareto distribution using inverse transform sampling \citep{newman_power_2005}. Lakes in my simulated dataset have a minimum and maximum area of approximately 1 and 81000 $km^2$ respectively. This maximum was chosen to be approximately as large as Lake Superior but less than the Caspian Sea following \citet{lehnerDevelopmentValidationGlobal2004}. The "true" total area of these lakes is approximately 230000 $km^2$. I simulated a censored lake dataset by excluding lakes smaller than $e^{1}$. This excludes (censors) approximately 60\% of the total dataset. I approximated the "true" lake area total by constructing the empirical distribution function (edf) of the data which approximates the underlying Pareto cdf \citep{newman_power_2005}. Then I used this estimate of the cdf slope to generate cdf estimates for the censored lakes. I combined these cdf estimates with the edf values from the known lakes before calculating the sum of the inverted distribution (Figure \ref{fig:predict_area}).

@@

Instead of the essentially fixed $a$ and total lake area estimates produced by the frequentist approach, I found a substantial variability in both $a$ (95\% CI: 0.86, 0.92) and total area using a Bayesian approach (Fig \ref{fig:bayesian_model}, \ref{fig:bayesian_area}). In particular, the Bayesian 95\% credible intervals for both $a$ and total lake area encapsulate the true values (Fig \ref{fig:bayesian_model}, \ref{fig:bayesian_area}). Despite better uncertainty estimates using a Bayesian approach, both the frequentist and Bayesian approaches underestimated the true value of $a$ and total lake area.

>>>>

\author{ \href{https://orcid.org/0000-0002-5924-2464}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}J.~Stachelek}\thanks{\href{https://jsta.rbind.io}{https://jsta.rbind.io}} \\
	Earth System Observations\\
	Los Alamos National Laboratory\\
	Los Alamos, NM, USA, 87544 \\
	\texttt{jsta@lanl.gov} \\
}

@@

\section{Introduction}
Area is a critical factor determining the contribution of lakes to global biogeochemical cycling. As such, existing evidence suggests that lakes are key components of overall carbon burial, sequestration, and greenhouse gas emissions \citep{delsontroGreenhouseGasEmissions2018, kellerGlobalCarbonBudget2021}. The calculation of such fluxes requires two terms 1) an areal flux rate and 2) the distribution of total lake area. As with all estimates at broad spatial scales, uncertainty estimates are of great interest both because they allow for intercomparisons among similar lake studies and because they allow for comparisons between lakes and other ecosystems.

The techniques typically used to determine uncertainty in areal flux rates and the techniques typically used to determine uncertainty in total lake area distributions differ greatly. In the former case, uncertainties can typically be calculated in straightforward manner from literature compilations \citep{delsontroGreenhouseGasEmissions2018, kellerGlobalCarbonBudget2021}. In the latter case, uncertainty in total lake area distributions is challenged by the fact that no existing database is a complete census of all lakes \citep{messagerEstimatingVolumeAge2016} and the distribution of individual lake areas spans a range of approximately 7 orders of magnitude. The largest lakes are so big they could otherwise be classified as inland seas (> $10^4$ $km^2$) while the smallest are barely larger than a regulation sized soccer (football) field (> $10^{-3}$ $km^2$).
A consequence of these challenges is that uncertainty varies within the distribution as a function of lake size. The area of the largest lakes is known exactly while the area of the smallest lakes is incomplete below a certain unknown threshold. This can occur either because lakes are too small to be resolved from remote sensing products or because of limited ground surveying effort. The omission or "censoring" of small lakes occurs because we know that small lakes exist but below a certain threshold we have limited knowledge of their exact areas \citep{hamiltonEstimationFractalDimension1992}.

Estimating total lake area from a sample of lakes requires a conceptual model of how lakes are formed (i.e. the data generating process). Typically, lake areas are treated as arising from a fractal generating process due to the fact that landform topography, which determines the placement of lakes, can itself be treated as a fractal generating process. Indeed, many other geomorphological phenomena that are dependent on landform topography such as coastline length are often well-described by fractal generating processes \citep{newman_power_2005}. A challenge in modelling such data generating processes for lake areas is that large lakes likely follow a different data generating process than that of the smallest lakes. Whereas small lakes are constrained by landform topography, large lakes are essentially unconstrained by local landform topography and are instead constrained by the placement and arrangement of continents \citep{goodchildLakesFractalSurfaces1988}. As large lakes become even larger, they have a greater probability of intersecting a continent edge and becoming an estuary or embayment rather than a lake. Consquently, lake databases are said to be truncated on large lakes because we know that large lakes are essentially fixed in space and cannot occur in any given location \citep{hamiltonEstimationFractalDimension1992}.

From the preceding discussion it is clear that estimating total lake area requires a method of dealing both with the fact that lake databases are 1) truncated at large lakes and 2) censored at small lakes. Previous studies estimating global lake area have essentially not dealt with the first issue but instead modified their estimation process to limit its effect on the results (see methods section). They have dealt with the second issue by specifying an ad-hoc cutoff value below which empirically determined lake areas are discarded and subsequently back calculated. Hereafter, I refer to this strategy as the "cutoff" method.

In the present study, I explore two different methods for calculating uncertainty bounds around global lake area. First, I calculate these bounds using a frequentist approach for estimating total lake area (i.e. the cutoff method) via a simulation study. Second, I compare these frequentist uncertainty bounds with those calculated using a Bayesian framework. I carry out these demonstrations using a simulated dataset so that the ability of each method to recover the "true" parameter values can be evaluated. The use of a simulated dataset has the further benefit of avoiding potential confounding factors such as heterogeneity of survey effort or unknown data precision.

@@

Lake areas are typically treated as arising from a scale-invariant fractal generating process \citep{winslowDoesLakeSize2015, downingGlobalAbundanceSize2006, mcdonald_regional_2012, goodchildLakesFractalSurfaces1988, hamiltonEstimationFractalDimension1992}. This means that the number of lakes in one size class is proportional to the number of lakes in the preceding size class irrespective of their magnitudes. The numerical form describing such a process is a power-law function. One of the statistical tools often used to model data that follow a power-law function is the Pareto distribution which has a probability density function (pdf) of:

@@

The reason for using the ccdf is two-fold. First, it stabilizes model estimates in the lower tail of the distribution \citep{newman_power_2005}. This can be seen from the simulated data in Figure \ref{fig:pareto_demo} where the Pareto pdf contains a lot of noise in the tail but the ccdf appears smoothed. The smoothing of the tail is also desirable because it functions as a way of dealing with the truncated nature of lake databases (i.e. the area of large lakes is known exactly). The second reason for using the ccdf is that it provides a computational shortcut for estimating the Pareto shape parameter $a$ because it is numerically equivalent to the slope of the ccdf in log-log space \citep{downingGlobalAbundanceSize2006}.

@@

For evaluation purposes, I generated a simulated dataset of 10,000 lake areas following the Pareto distribution using inverse transform sampling \citep{newman_power_2005}. Lakes in my simulated dataset have a minimum and maximum area of approximately 1 and 81000 $km^2$ respectively. This maximum was chosen to be approximately as large as Lake Superior but less than the Caspian Sea following \citet{lehnerDevelopmentValidationGlobal2004}. The "true" total area of these lakes is approximately 230000 $km^2$. I simulated a censored lake dataset by excluding lakes smaller than $e^{1}$. This excludes (i.e. censors) approximately 60\% of the total dataset. I approximated the "true" lake area total by constructing the empirical distribution function (edf) of the data which approximates the underlying Pareto cdf \citep{newman_power_2005}. Then I used this estimate of the cdf slope to generate cdf estimates for the censored lakes. I combined these cdf estimates with the edf values from the known lakes before calculating the sum of the inverted distribution (Figure \ref{fig:predict_area}).

@@

Instead of the essentially fixed $a$ and total lake area estimates produced by the frequentist approach, I found substantial variability in both $a$ (95\% CI: 0.86, 0.92) and total area using a Bayesian approach (Fig \ref{fig:bayesian_model}, \ref{fig:bayesian_area}). In particular, the Bayesian 95\% credible intervals for both $a$ and total lake area encapsulate the true values (Fig \ref{fig:bayesian_model}, \ref{fig:bayesian_area}). Despite better uncertainty estimates using a Bayesian approach, both the frequentist and Bayesian approaches underestimated the true value of $a$ and total lake area.
